### 2025-07-19 docs/article/Sections/results.tex
Updated results section

However, it is important to acknowledge the limitations of our chosen data
models for generalizability. The tested templates used relatively simple data
structures with limited nesting and straightforward property bindings.
Real-world applications often involve deeply nested data structures, complex
conditional logic, and iterative rendering over thousands of items. Under such
conditions, we anticipate significantly different performance characteristics:
increased memory consumption due to object traversal overhead, higher CPU
utilization for complex evaluations, and altered scalability patterns where
non-blocking approaches may become more advantageous. The performance
degradation observed with our Stock class benchmark—which merely doubled the
number of properties—suggests that enterprise-level template complexity would
result in substantially higher performance degradation. Future work should
investigate these scenarios with more realistic data models to better
understand performance boundaries and optimization strategies for complex PSSR
implementations.

@@

This section evaluates the memory and CPU resource characteristics of three
representative concurrency models: Virtual Threads, suspendable coroutines, and
reactive streams. While previous benchmarks demonstrated that Virtual Threads
can offer comparable throughput to non-blocking paradigms, we now assess
whether this simplicity in code comes with higher runtime resource costs.

Resource utilization data was collected using VisualVM during 40-second
benchmark runs, capturing detailed CPU and memory usage patterns under
sustained load conditions. Each profiling session monitored system performance
during the 1 to 128 concurrent user load test, providing comprehensive insights
into runtime behavior.

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
     \caption{CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:cpu-virt}
\end{figure}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-susp.png}
     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:cpu-susp}
\end{figure}

Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
respectively. In both cases, the initial spike in CPU usage corresponds to the
system's bootstrap and initial load phase. The subsequent gradual increase in
CPU utilization reflects the ramp-up of concurrent users during the benchmark.
\texttt{HtmlFlow-Virtual} exhibits a distinctive bell-shaped curve, rapidly
climbing from near-zero to peak CPU usage of around 90\% during the initial
load phase, maintaining high utilization during sustained load, and then
dropping sharply back to baseline. The profile shows more aggressive CPU
consumption with steep ramp-up and ramp-down phases. In contrast,
\texttt{HtmlFlow-Susp} displays a more gradual and controlled resource
utilization pattern, with a smoother ascent to a lower peak of approximately
75\%, followed by a more gentle descent. This smoother profile suggests more
efficient CPU scheduling and better resource management under load. The GC
activity indicators in both profiles remain relatively low throughout the test
duration, confirming that garbage collection overhead does not significantly
impact CPU utilization in either approach.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
     \caption{Memory utilization and garbage collection behavior with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:gc-virt}
\end{figure}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:gc-susp}
\end{figure}

The memory usage disparity is more pronounced. Memory utilization data was
collected using VisualVM profiling during the same 40-second benchmark runs. As
shown in Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually
climbs to a peak heap usage of 900MB, with large but infrequent garbage
collection events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks
at only 245MB, following a regular sawtooth pattern of small, frequent garbage
collections. Overall, \texttt{HtmlFlow-Virtual} allocates a total of 939MB heap
compared to just 348MB in \texttt{HtmlFlow-Susp}—a 2.7× difference.

The higher observed heap usage with Virtual Threads stems from how they manage
concurrency. Virtual Threads allocate a full stack per thread, and when they
are suspended (e.g., during blocking I/O), their entire call stack is unmounted
and copied to the heap, allowing the physical carrier thread to be reused.
Kotlin coroutines, by contrast, also store their execution state on the heap
when suspended — but the compiler transforms coroutine code into a state
machine, so only the minimal necessary state is saved rather than a full stack.
This results in significantly lower memory overhead per suspended task.

It is important to note that our benchmark keeps data items in memory as static
objects, which represents a best-case scenario for memory allocation patterns.
In real-world applications, data typically comes from external sources such as
databases, web services, or file systems, requiring dynamic object creation and
serialization for each request. This would result in significantly higher
memory allocation rates and more frequent garbage collection cycles across all
concurrency models, with the observed differences between Virtual Threads and
suspendable coroutines likely becoming even more pronounced under such
realistic I/O-intensive workloads.

\subsubsection{Summary of Trade-offs}

While the approaches can achieve similar throughput under load, they differ
significantly in runtime efficiency. Virtual Threads, despite offering a
simplified synchronous style, incur the highest memory and CPU overhead.
structured concurrency delivers superior resource efficiency but
requires familiarity with asynchronous programming constructs. The RxJava
blocking iterator, while easier to adopt in traditional reactive settings,
performs worse in memory-constrained environments due to its unbounded internal
queuing model.

These results suggest that system resource constraints and predictability
requirements should inform the choice of concurrency model. In environments
where memory and CPU resources are limited, structured concurrency via
suspendable coroutines is the most efficient and robust option.

>>>>

However, it is important to acknowledge the limitations of our chosen data
models for generalizability. The tested templates used relatively simple data
structures with limited nesting and straightforward property bindings.
Real-world applications often involve deeply nested data structures, complex
conditional logic, and iterative rendering over thousands of items. Under such
conditions, we anticipate significantly different performance characteristics:
increased memory consumption due to object traversal overhead, higher CPU
utilization for complex evaluations, and altered scalability patterns where
non-blocking approaches may become more advantageous. The performance
degradation observed with our Stock class benchmark—which merely doubled the
number of properties—suggests that higher template complexity would result in
substantially higher performance degradation. Future work should investigate
these scenarios with more realistic data models to better understand
performance boundaries and optimization strategies for complex PSSR
implementations.

@@

This section evaluates the memory and CPU resource usage characteristics of
structured concurrency approaches, specifically comparing Virtual Threads and
suspendable coroutines.

Resource utilization data was collected using VisualVM during 30-second
benchmark runs, capturing detailed CPU and memory usage patterns under
sustained load conditions. Each profiling session monitored system performance
during the 1 to 128 concurrent user load test, providing comprehensive insights
into runtime behavior.

@@

\begin{center}
     \begin{minipage}[t]{0.48\textwidth}
          \centering
          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-susp.png}
          \captionof{figure}{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
          \label{fig:cpu-susp}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{0.48\textwidth}
          \centering
          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-virt.png}
          \captionof{figure}{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
          \label{fig:cpu-virt}
     \end{minipage}
\end{center}

\begin{center}
     \begin{minipage}[t]{0.48\textwidth}
          \centering
          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-susp.png}
          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
          \label{fig:gc-susp}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{0.48\textwidth}
          \centering
          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-virt.png}
          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
          \label{fig:gc-virt}
     \end{minipage}
\end{center}

Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
respectively. The subsequent gradual increase in CPU utilization reflects the
ramp-up of concurrent users during the benchmark. \texttt{HtmlFlow-Virtual}
exhibits a distinctive bell-shaped curve, rapidly climbing from near-zero to
peak CPU usage of around 50\% during the sustained load phase, maintaining
moderate utilization during the test period, and then dropping sharply back to
baseline. \texttt{HtmlFlow-Susp} displays a similar pattern but with a slightly
lower peak CPU utilization of approximately 42\%, showing marginally lower
intensive CPU usage during the sustained load phase. Both profiles demonstrate
comparable resource consumption patterns, with the virtual thread
implementation showing only slightly higher CPU demands. The GC activity
indicators in both profiles remain relatively low throughout the test duration,
confirming that garbage collection overhead does not significantly impact CPU
utilization in either approach.

Figures~\ref{fig:gc-susp} and~\ref{fig:gc-virt} show similar memory usage for
both approaches, with peaks around 300~MB during load. The initial spike to
450~MB in both graphs corresponds to the applications' bootstrapping phase.
Overall, these results indicate that under the tested conditions, Virtual
Threads and suspendable coroutines exhibit comparable memory consumption.

Virtual Threads allocate a full call stack per thread, which is unmounted and
stored on the heap when suspended (e.g., during blocking I/O). In contrast,
Kotlin coroutines compile into state machines that capture only the minimal
execution state needed to resume computation. This typically allows coroutines
to scale more efficiently in terms of memory usage, especially under high
concurrency. These findings contrasts with prior research, such as Beronić et
al.~\cite{9803765}, which reported lower heap usage for Kotlin coroutines
(52--99~MB) compared to Java Virtual Threads (16--64~MB), albeit under
different workloads.

These observations suggest that the choice between Virtual Threads and
suspendable coroutines should be guided by both runtime behavior and
development preferences. Virtual Threads offer the simplicity of synchronous
programming with competitive performance in moderate-load scenarios. In
contrast, suspendable coroutines, while requiring an asynchronous programming
model, may offer better efficiency in applications with higher concurrency,
deeper stacks, or more dynamic workloads.
